Fixed-Parameter Tractable Sampling for RNA Design with Multiple Target Structures

### 2019-01-24 RNARedPrint.tex

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$, associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample from the Boltzmann distribution, defined over the set of sequences that satisfy classic base pairing rules for each structure as: 
\todo{$S$ sequence not defined!}
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences:
%Given are the features $F_1,\dots,F_k$, which encode the energies functions for the fixed structures and the function for the \GCb\%, all in terms of simple atomic energy contributions.

For weights $\pi_1,\dots,\pi_k$, one wants to compute the partition function
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)}.
  \end{equation}

\subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
\todo{$x_1$ position(?) not defined!\\$S_x_1$ nucleotide at position(?) not defined!\\$p$ number of dependent positions(?) from $x_p$ not defined!}
Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}. Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
$$
F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,\ldots,x_p\}$}}}\!\!\!\!\!\! f\left(\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}\right).
$$

The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, $i\in[1,n]$, whose dependencies are restricted to individual nucleotides,  and such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$ 
\todo{Why here $i$ and not $x_i$??}
\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence, one simply counts the occurrences of $\Gb$ and $\Cb$.}

>>>>

%\paragraph{Central computational problem.}
Let us consider $m$ (secondary) structures $R_1,\dots, R_m$, each abstracted as a set of base pairs, and \revised{$k\geq m$} features $F_1,\dots, F_k$\revised{, typically representing the energies of the structures and additional sequence properties,} associated with weights $\pi_1,\dots,\pi_k$. Our goal is to sample sequences \revised{$S$ (which satisfy the base pairing rules for all strucutures) from the Boltzmann distribution defined by} 
\begin{equation}
\label{eq:sample-distribution}
\Pr(S\mid\pi_1,\dots,\pi_k) \propto \displaystyle\prod_{1\leq \ell\leq k}\!\! \pi_\ell^{-F_\ell(S)} \end{equation}

The workhorse of our approach is the \revised{fixed-parameter tractable} computation of feature-dependent partition functions over sequences, \revised{namely partition functions of the form} 
%
  \begin{equation}
    \label{eq:mainproblem}
    \partfun{\pi_1,\dots,\pi_k} = \sum_{S\in\Sigma^n} \prod_{1\leq \ell\leq k} \pi_\ell^{-F_\ell(S)},
  \end{equation}
  for specfic weights $\pi_1,\dots,\pi_k$.

\subsection*{Expressing \GCb\%-content, sequence validity and energies as features}
Formally, we define a \Def{feature $F$} as a function on sequences, whose value is obtained by summing over an associated set of \Def{contributions}. Each contribution $f$ takes values in $\mathbb{R}\cup \{+\infty\}$, and depends on the nucleotides assigned to a restricted set of positions, its \emph{dependencies $\dep(f)$}, such that
}
$$
F(S) =\!\!\!\!\!\!\!\sum_{\substack{\text{$f$ contribution of $F$,}\\\text{ $\dep(f)=\{x_1,x_2,\ldots\}$}}}\!\!\!\!\!\! f\left(\left\{\substack{x_1\mapsto S_{x_1}\\x_p\mapsto S_{x_2}\\\cdots}\right\}\right).
$$
\revised{Here, since $\dep(f)=\{x_1,\ldots,x_p\}$, $\left\{\substack{x_1\mapsto S_{x_1}\\\cdots\\x_p\mapsto S_{x_p}}\right\}$ denotes the assignment, that assigns to all positions $x_k$  in $\dep(f)$ the respective nucleotide $S_{x_k}$ ($1\leq k\leq p$).}

The $\GCb\%$ can be simply expressed using $n$ contributions $f^{\GCb}_i$, \revised{each dependencing only on position $i\in[1,n]$, i.e.~$\dep(f^{\GCb}_i)=\{i\}$}, such that 
$$f^{\GCb}_i(\{i\mapsto c\})=\begin{cases}-1&\text{ if }c=\Gb\text{ or }\Cb\\ 0&\text{ otherwise}.\end{cases}$$

\revised{By summing $f^{\GCb}_i(\{i\mapsto S_i\})$ over the whole sequence ($i=1,\dots,n$), one simply counts the occurrences of $\Gb$ and $\Cb$.}
